---
title: "Arctic FRED Analysis"
format: html

---

```{r, include = FALSE}
library(mosaic)
library(ggformula)
library(tidyverse)
library(stringr)
library(janitor)
library(sf)
library(geosphere)
library(fuzzyjoin)
library(dplyr)
library(rnaturalearth)
library(ggnewscale)

options(digits = 6)
```

# Importing Data

```{r}
# Importing data sets
FRED3 <- read.csv("FRED3.csv")
FRED4 <- read.csv("FRED4.csv")
FRED_newest <- readr::read_csv("FRED_NEWEST.csv", 
                        locale = readr::locale(encoding = "latin1"))

AU <- read.csv("AUwarming.csv")
```

# Cleaning Data

## FRED 3.0 Cleaning

```{r, warning = FALSE}
# FRED 3 data cleaning

# make first row the header
FRED3 <- FRED3 |> row_to_names(1)

# remove irrelevant rows
FRED3 <- FRED3 |>
  filter(Name == "1" | Name == "Type")

# pulling the first row as a vector 
var_type <- unlist(FRED3[1,], use.names = FALSE)

# get vector of indices that correspond to root traits
indices <- rep(0, length(var_type))

# looping through entire row of variable types
for (i in 1:length(var_type)) {
  
  # if entry is empty
  if(is.na(var_type[i])){
    # skip that entry
    i = i + 1
  }
  # if the variable type is root trait
  else if (var_type[i] == "Root_Traits") {
   
    # add the index to our vector
    indices[i] <- i

  }
}

# dropping "0" values from indices vector
indices <- indices[indices != 0]

# dplyr::selecting columns with article, lat, long, and root traits
FRED3_cleaned <- FRED3 |>
  # create short variable name for article
  rename(Article = "Abbreviated article citation", 
         order = "Plant taxonomy_Accepted order_APW",
         family = "Plant taxonomy_Accepted family_TPL", 
         genus = "Plant taxonomy_Accepted genus_TPL", 
         species = "Plant Taxonomy_Accepted species_TPL") |>
  # select article, coordinates, and root trait columns
  dplyr::select(Article, Latitude_Main, Longitude_Main,
                order, family, genus, species, indices) 
 
# removing the row that was "type"
FRED3_cleaned <- FRED3_cleaned[3:nrow(FRED3_cleaned),]

# keeping only non-empty columns
FRED3_cleaned <- FRED3_cleaned |>
  select(where(~ any(!is.na(.) & . != "")))


```

## FRED 4.0 Cleaning


```{r, warning = FALSE}
# FRED 4 data cleaning

# making first row replace the header
FRED4 <- FRED4 |> row_to_names(1)

# remove irrelevant rows
FRED4 <- FRED4 |>
  filter(`Data_Raw or unpublished` == "main article" | Name == "Type")

# pulling the first row as a vector 
var_type <- unlist(FRED4[1,], use.names = FALSE)

# get vector of indices that correspond to root traits
indices <- rep(0, length(var_type))

# looping through entire row of variable types
for (i in 1:length(var_type)) {
  
  # if entry is empty
  if(is.na(var_type[i])){
    # skip that entry
    i = i + 1
  } 
  # if the variable type is root trait
  else if (var_type[i] == "Root_Traits") {
   
    # add the index to our vector
    indices[i] <- i

  }
}

# dropping "0" values from indices vector
indices <- indices[indices != 0]

# selecting columns with article, lat, long, and root traits
FRED4_cleaned <- FRED4 |>
   # create short variable name for article
  rename(Article = "Abbreviated article citation",
         citation = "Data source_Citation",
         order = "Plant taxonomy_Accepted order_APW",
         family = "Plant taxonomy_Accepted family_TPL", 
         genus = "Plant taxonomy_Accepted genus_TPL", 
         species = "Plant Taxonomy_Accepted species_TPL") |>
  # select article, coordinates, and root trait columns
  dplyr::select(Article, citation, Latitude_Main, Longitude_Main, 
                order, family, genus, species, indices) 

# removing the row that represented "type"
FRED4_cleaned <- FRED4_cleaned[3:nrow(FRED4_cleaned),]

# keeping only non-empty columns
FRED4_cleaned <- FRED4_cleaned |>
  select(where(~ any(!is.na(.) & . != "")))

```

## Arctic FRED Cleaning

```{r}

# making first row replace the header
FRED4new <- FRED_newest |> row_to_names(1)

# remove irrelevant rows
FRED4new <- FRED4new |>
  filter(`Data_Raw or unpublished` == "main article" |
           `Data_Raw or unpublished` == "supplement" | 
           `Data source_Citation` == "Meta_Data")

# pulling the first row as a vector 
var_type <- unlist(FRED4new[1,], use.names = FALSE)

# get vector of indices that correspond to root traits
indices <- rep(0, length(var_type))

# looping through entire row of variable types
for (i in 1:length(var_type)) {
  
  # if entry is empty
  if(is.na(var_type[i])){
    # skip that entry
    i = i + 1
  }
  # if the variable type is root trait
  else if (var_type[i] == "Root_Traits") {
   
    # add the index to our vector
    indices[i] <- i

  }
}

# dropping "0" values from indices vector
indices <- indices[indices != 0]

# selecting columns with article, lat, long, and root traits
FRED4new_cleaned <- FRED4new |>
   # create short variable name for article
  rename(Article = "Abbreviated article citation", 
         citation = "Data source_Citation",
         order = "Plant taxonomy_Accepted order_WFO",
         family = "Plant taxonomy_Accepted family_WFO", 
         genus = "Plant taxonomy_Accepted genus_WFO", 
         species = "Plant Taxonomy_Accepted species_WFO") |>
  # select article, coordinates, and root trait columns
  dplyr::select(Article, citation, Latitude_Main, 
                order, family, genus, species, 
                Longitude_Main, indices)

# removing the row that represented "type"
FRED4new_cleaned <- FRED4new_cleaned[3:nrow(FRED4new_cleaned),]

# # keeping only non-empty columns
FRED4new_cleaned <- FRED4new_cleaned |>
  select(where(~ any(!is.na(.) & . != "")))

```


## Stats about the FRED 3, 4.0 and 4.1

### FRED 3.0 Arctic Stats

```{r}
FRED3_arctic <- FRED3_cleaned |>
  mutate(Latitude_Main = as.numeric(Latitude_Main),
         Longitude_Main = as.numeric(Longitude_Main)) |>
  filter(
    # Global baseline: Everything north of 66.5333
    (Latitude_Main >= 66.5333 & Latitude_Main <= 90.0) |
    
    # North America exception: North of 60°N between 20W and 170W
    (Latitude_Main >= 60.0 & Latitude_Main < 66.5333 & 
       Longitude_Main >= -170 & Longitude_Main <= -20) |
    
    # Asia exception: North of 62N between 60E and 180E 
    (Latitude_Main >= 62.0 & Latitude_Main < 66.5333 & 
       ((Longitude_Main >= 60 & Longitude_Main <= 180) | 
        (Longitude_Main >= -180 & Longitude_Main <= -170)))
  )

FRED3_arctic |>
  distinct(Article) |>
  nrow()

FRED3_unique_species <- FRED3_arctic |>
  select(genus, species) |>
  distinct(genus, species, .keep_all = TRUE) |>
  mutate(across(everything(), ~ na_if(str_trim(.x), ""))) |>
  distinct(genus, species, .keep_all = TRUE) |>
  remove_empty(which = "rows") |>
  nrow()

FRED3_unique_species

FRED3_unique_coord <- FRED3_arctic |>
  select(Latitude_Main, Longitude_Main) |>
  drop_na(Latitude_Main, Longitude_Main) |>
  filter(
    Latitude_Main != "unspecified",
    Longitude_Main != "unspecified"
  ) |>
  distinct() |>
  nrow()

FRED3_unique_coord

FRED3arctic_sf <- st_as_sf(
  FRED3_arctic,
  coords = c("Longitude_Main", "Latitude_Main"),
  crs = 4326,
  remove = FALSE # Keeps  original coordinate columns in the output
)

# get the world map data
countries_sf <- ne_countries(scale = "medium", returnclass = "sf")

# spatial join to find which country polygon each point sits inside
# adds 'country_name' and 'country_iso' to  data
FRED3arctic_with_countries <- st_join(
  FRED3arctic_sf,
  countries_sf[, c("name", "iso_a3")]
)

# Rename the new columns for clarity
names(FRED3arctic_with_countries)[names(FRED3arctic_with_countries) == "name"] <- "Country_Name"
names(FRED3arctic_with_countries)[names(FRED3arctic_with_countries) == "iso_a3"] <- "Country_ISO"

# convert back to a regular data frame/tibble
FRED3arctic_with_countries <- as.data.frame(FRED3arctic_with_countries)

range(FRED3arctic_with_countries$Latitude_Main, na.rm = TRUE)
range(FRED3arctic_with_countries$Longitude_Main, na.rm = TRUE)

FRED3arctic_country <- FRED3arctic_with_countries |>
  select(Latitude_Main, Longitude_Main, Country_Name) |>
  drop_na(Country_Name) |>
  distinct(Country_Name)

FRED3arctic_country$Country_Name

FRED3arctic_countryyr <- FRED3arctic_with_countries |>
  mutate(year = str_extract(Article, "[0-9]{4}"))

range(FRED3arctic_countryyr$year, na.rm = TRUE)

FRED3distinctyr <- FRED3arctic_countryyr |>
  distinct(year)

count(FRED3distinctyr)

# distinct(Article)
```

### FRED 4.0 Stats

```{r}

FRED4_arctic <- FRED4_cleaned |>
  mutate(Latitude_Main = as.numeric(Latitude_Main),
         Longitude_Main = as.numeric(Longitude_Main)) |>
  filter(
    # Global baseline: Everything north of 66.5333
    (Latitude_Main >= 66.5333 & Latitude_Main <= 90.0) |
    
    # North America exception: North of 60°N between 20W and 170W
    (Latitude_Main >= 60.0 & Latitude_Main < 66.5333 & 
       Longitude_Main >= -170 & Longitude_Main <= -20) |
    
    # Asia exception: North of 62N between 60E and 180E 
    (Latitude_Main >= 62.0 & Latitude_Main < 66.5333 & 
       ((Longitude_Main >= 60 & Longitude_Main <= 180) | 
        (Longitude_Main >= -180 & Longitude_Main <= -170)))
  )

FRED4_arctic |>
  distinct(Article) |>
  nrow()

FRED4_unique_species <- FRED4_arctic |>
  select(genus, species) |>
  distinct(genus, species, .keep_all = TRUE) |>
  mutate(across(everything(), ~ na_if(str_trim(.x), ""))) |>
  distinct(genus, species, .keep_all = TRUE) |>
  remove_empty(which = "rows") |>
  nrow()

FRED4_unique_species

FRED4_unique_coord <- FRED4_arctic |>
  select(Latitude_Main, Longitude_Main) |>
  drop_na(Latitude_Main, Longitude_Main) |>
  filter(
    Latitude_Main != "unspecified",
    Longitude_Main != "unspecified"
  ) |>
  distinct() |>
  nrow()

FRED4_unique_coord

FRED4arctic_sf <- st_as_sf(
  FRED4_arctic,
  coords = c("Longitude_Main", "Latitude_Main"),
  crs = 4326,
  remove = FALSE # Keeps  original coordinate columns in the output
)

# get the world map data
countries_sf <- ne_countries(scale = "medium", returnclass = "sf")

# spatial join to find which country polygon each point sits inside
# adds 'country_name' and 'country_iso' to  data
FRED4arctic_with_countries <- st_join(
  FRED4arctic_sf,
  countries_sf[, c("name", "iso_a3")]
)

# Rename the new columns for clarity
names(FRED4arctic_with_countries)[names(FRED4arctic_with_countries) == "name"] <- "Country_Name"
names(FRED4arctic_with_countries)[names(FRED4arctic_with_countries) == "iso_a3"] <- "Country_ISO"

# convert back to a regular data frame/tibble
FRED4arctic_with_countries <- as.data.frame(FRED4arctic_with_countries)

range(FRED4arctic_with_countries$Latitude_Main, na.rm = TRUE)
range(FRED4arctic_with_countries$Longitude_Main, na.rm = TRUE)

FRED4arctic_country <- FRED4arctic_with_countries |>
  select(Latitude_Main, Longitude_Main, Country_Name) |>
  drop_na(Country_Name) |>
  distinct(Country_Name)

FRED4arctic_country$Country_Name

FRED4arctic_countryyr <- FRED4arctic_with_countries |>
  mutate(year = str_extract(Article, "[0-9]{4}"))

range(FRED4arctic_countryyr$year, na.rm = TRUE)

FRED4distinctyr <- FRED4arctic_countryyr |>
  distinct(year)

count(FRED4distinctyr)
```
### FRED 4.1 Stats

```{r}

FRED4new_arctic <- FRED4new_cleaned |>
  mutate(Latitude_Main = as.numeric(Latitude_Main),
         Longitude_Main = as.numeric(Longitude_Main)) |>
  filter(
    # Global baseline: Everything north of 66.5333
    (Latitude_Main >= 66.5333 & Latitude_Main <= 90.0) |
    
    # North America exception: North of 60°N between 20W and 170W
    (Latitude_Main >= 60.0 & Latitude_Main < 66.5333 & 
       Longitude_Main >= -170 & Longitude_Main <= -20) |
    
    # Asia exception: North of 62N between 60E and 180E 
    (Latitude_Main >= 62.0 & Latitude_Main < 66.5333 & 
       ((Longitude_Main >= 60 & Longitude_Main <= 180) | 
        (Longitude_Main >= -180 & Longitude_Main <= -170)))
  )

FRED4new_arctic |>
  distinct(Article) |>
  nrow()

FRED4new_unique_species <- FRED4new_arctic |>
  select(genus, species) |>
  distinct(genus, species, .keep_all = TRUE) |>
  mutate(across(everything(), ~ na_if(str_trim(.x), ""))) |>
  distinct(genus, species, .keep_all = TRUE) |>
  remove_empty(which = "rows") |>
  nrow()

FRED4new_unique_species

FRED4new_unique_coord <- FRED4new_arctic |>
  select(Latitude_Main, Longitude_Main) |>
  drop_na(Latitude_Main, Longitude_Main) |>
  filter(
    Latitude_Main != "unspecified",
    Longitude_Main != "unspecified"
  ) |>
  distinct() |>
  nrow()

FRED4new_unique_coord

FRED4newarctic_sf <- st_as_sf(
  FRED4new_arctic,
  coords = c("Longitude_Main", "Latitude_Main"),
  crs = 4326,
  remove = FALSE # Keeps  original coordinate columns in the output
)

# get the world map data
countries_sf <- ne_countries(scale = "medium", returnclass = "sf")

# spatial join to find which country polygon each point sits inside
# adds 'country_name' and 'country_iso' to  data
FRED4newarctic_with_countries <- st_join(
  FRED4newarctic_sf,
  countries_sf[, c("name", "iso_a3")]
)

# Rename the new columns for clarity
names(FRED4newarctic_with_countries)[names(FRED4newarctic_with_countries) == "name"] <- "Country_Name"
names(FRED4newarctic_with_countries)[names(FRED4newarctic_with_countries) == "iso_a3"] <- "Country_ISO"

# convert back to a regular data frame/tibble
FRED4newarctic_with_countries <- as.data.frame(FRED4newarctic_with_countries)

range(FRED4newarctic_with_countries$Latitude_Main, na.rm = TRUE)
range(FRED4newarctic_with_countries$Longitude_Main, na.rm = TRUE)

FRED4newarctic_country <- FRED4newarctic_with_countries |>
  select(Latitude_Main, Longitude_Main, Country_Name) |>
  drop_na(Country_Name) |>
  distinct(Country_Name)

FRED4newarctic_country$Country_Name

FRED4newarctic_countryyr <- FRED4newarctic_with_countries |>
  mutate(year = str_extract(Article, "[0-9]{4}"))

range(FRED4newarctic_countryyr$year, na.rm = TRUE)

FRED4newdistinctyr <- FRED4newarctic_countryyr |>
  distinct(year)

FRED4newdistinctyr

count(FRED4newdistinctyr)
```


## Combining FRED 4.0 and FRED 4.1 with Warming Column

```{r}
clean_title <- function(x) {
  x |>
    as.character() |>
    iconv(from = "UTF-8", to = "ASCII//TRANSLIT", sub = "") |>
    stringr::str_to_lower() |>
    stringr::str_replace_all("[[:punct:]]", " ") |>
    stringr::str_squish()
}

AU_clean <- AU |>
  mutate(title_clean = clean_title(Article.Title)) |>
  rename(warming = "Should.be.in.Warming.dataset...y.n.") |>
  dplyr::select(title_clean, warming)

FRED4_titles <- FRED4_cleaned |>
  mutate(F00004_clean = clean_title(citation))

FRED4_ifwarming <- fuzzy_left_join(
  FRED4_titles,
  AU_clean,
  by = c("F00004_clean" = "title_clean"),
  match_fun = ~ str_detect(.x, fixed(.y))
) |>
  select(-F00004_clean, -title_clean)

FRED4new_titles <- FRED4new_cleaned |>
  mutate(F00004_clean = clean_title(citation))

FRED4new_ifwarming <- fuzzy_left_join(
  FRED4new_titles,
  AU_clean,
  by = c("F00004_clean" = "title_clean"),
  match_fun = ~ str_detect(.x, fixed(.y))
) |>
  select(-F00004_clean, -title_clean)

# table(FRED4_titles$Article)
# table(FRED4_warming$Article)
# table(FRED4new_titles$Article)
# table(FRED4new_warming$Article)
# 
# ncol(FRED4_warming)
# ncol(FRED4new_warming)
```

# Graphs

## Full FRED Map

```{r}

FRED3_arctic <- FRED3_arctic |>
  mutate(warming = "n")

FRED4_combined <- bind_rows(
  FRED4_ifwarming,
  FRED4new_ifwarming)

fullfred_arctic <- bind_rows(
  FRED3_arctic,
  FRED4_combined |>
    mutate(
      Latitude_Main = as.numeric(Latitude_Main),
      Longitude_Main = as.numeric(Longitude_Main)
    ) |>
    filter(
      (Latitude_Main >= 66.5333 & Latitude_Main <= 90) |
      (Latitude_Main >= 60 & Latitude_Main < 66.5333 &
         Longitude_Main >= -170 & Longitude_Main <= -20) |
      (Latitude_Main >= 62 & Latitude_Main < 66.5333 &
         ((Longitude_Main >= 60 & Longitude_Main <= 180) |
          (Longitude_Main >= -180 & Longitude_Main <= -170)))
    )
)

fullfred_map <- fullfred_arctic |>
  distinct(Article, .keep_all = TRUE) |>
  mutate(
    Latitude_Main = as.numeric(Latitude_Main),
    Longitude_Main = as.numeric(Longitude_Main),
    warming = replace_na(warming, "n")
  ) |>
  filter(
    !is.na(Latitude_Main),
    !is.na(Longitude_Main)
  ) |>
  mutate(
    Category = case_when(
      warming == "y" ~ "Warming and Arctic",
      TRUE ~ "Arctic"
    )
  )

sites_sf <- st_as_sf(
  fullfred_map,
  coords = c("Longitude_Main", "Latitude_Main"),
  crs = 4326,
  remove = FALSE
)

world <- ne_countries(
  scale = "medium",
  returnclass = "sf"
)

sites_country <- st_join(
  sites_sf,
  world[, "name"]
)

country_counts <- sites_country |>
  st_drop_geometry() |>
  distinct(Article, name) |>
  count(name, name = "n_articles")

world_counts <- world |>
  left_join(country_counts, by = "name")

world_counts$n_articles[is.na(world_counts$n_articles)] <- 0

arctic_crs <- 3995

world_arctic <- st_transform(world_counts, arctic_crs)
sites_arctic <- st_transform(sites_country, arctic_crs)

ggplot() +
  # Base Map Layer (Continuous Fill)
  geom_sf(
    data = world_arctic,
    aes(fill = n_articles),
    color = "grey40",
    linewidth = 0.1
  ) +
  scale_fill_distiller(
    palette = "Purples",
    direction = 1,
    name = "Articles"
  ) +
  
  # Clear fill scale for next layer
  new_scale_fill() +
  
  # Points Layer (Discrete Fill with Black Borders)
  geom_sf(
    data = sites_arctic,
    aes(shape = Category, fill = Category), 
    colour = "black",                      
    size = 2,
    stroke = 0.8                           
  ) +
  scale_shape_manual(
    values = c(
      Arctic = 24,              # Filled Triangle
      Warming = 23,             # Filled Diamond
      "Warming and Arctic" = 21 # Filled Circle
    )
  ) +
  scale_fill_manual(            # This now safely applies only to the points!
    values = c(
      Arctic = "#00A5CF",
      "Warming and Arctic" = "#D7263D",
      Warming = "orange"
    )
  ) +
  coord_sf(
    xlim = c(-4e6, 4e6),
    ylim = c(-4e6, 4e6)
  ) +
  theme_minimal() +
  labs(
    shape = "Study Type",
    fill = "Study Type"
  )

# ggsave(
#   filename = "arctic_warming_map.png",
#   width = 8,
#   height = 8,
#   units = "in",
#   dpi = 300
# )
```


## Articles Per Year

```{r}

# aggregate data by Year and Category
yearly_trends <- fullfred_arctic |>
  # Ensure we only count each article once 
  distinct(Article, .keep_all = TRUE) |>
  # Clean up and extract the 4-digit year from the citation
  mutate(
    Year = as.numeric(str_extract(Article, "[0-9]{4}")),
    Category = case_when(
      warming == "y" ~ "Warming and Arctic",
      TRUE ~ "Arctic Only"
    )
  ) |>
  # Drop any rows where the year couldn't be parsed
  # drop_na(Year) |>
  # Count the number of unique papers per year per category
  group_by(Year, Category) |>
  summarise(Paper_Count = n(), .groups = "drop")

# plot the timeline trend
ggplot(yearly_trends, aes(x = Year, y = Paper_Count, fill = Category)) +
  # Using geom_col() to make a clean, non-overlapping stacked timeline bar chart
  geom_col(width = 1, color = "black", linewidth = 0.2) + 
  scale_fill_manual(
    values = c(
      "Arctic Only" = "#00A5CF",         
      "Warming and Arctic" = "#D7263D"  
    )
  ) +
  # Expand scale to catch the full timeline (1931 - 2023)
  scale_x_continuous(breaks = seq(1930, 2025, by = 10)) +
  scale_y_continuous(expand = c(0, 0), # Rest bars directly on the x-axis
                     breaks = seq(0, 
                                  max(yearly_trends$Paper_Count, na.rm = TRUE) + 2, by = 2)) + 
  theme_minimal() +
  labs(
    x = "Publication Year",
    y = "Number of Papers",
    fill = "Study Type"
  ) +
  theme(
    # panel.grid.minor = element_blank(),
    # panel.grid.major.x = element_blank(), # Clears vertical lines for an elegant look
    text = element_text(size = 11),
    legend.position = "bottom"            # Places legend nicely beneath the plot
  )



```

## Root traits measured

Most commonly measured
Root traits per article included in database, and average
Observations per article, and avg

```{r, fig.height=15, fig.width=9}
fullfred_arctic_pivot <- fullfred_arctic |> 
  # select non-empty columns
  dplyr::select(-Latitude_Main, -Longitude_Main, -warming,
                -order, -family, -genus, -species, -Elevation) |>
  # keep columns article, lat, long and only pivot root trait columns (#4-738)
  pivot_longer(cols = 2:743,
               names_to = "Trait",
               values_to = "Value",
               values_drop_na = TRUE) |>
  # turn all empty cells into NAs
  mutate_all(na_if, "") |>
  # remove NAs
  drop_na()

# filter unique traits per article 
# count number of traits and number of articles they are in 
countsbyrt <- fullfred_arctic_pivot |>
  # drop non-numeric value
  mutate(as.numeric(Value)) |>
  drop_na() |>
  group_by(Article) |>
  # list unique root traits for each article
  summarize(Type = unique(Trait)) |>
  # add year column
  mutate(Year = str_extract(Article, "[0-9]{4}")) |>
  group_by(Type) |>
  # count number of times the root trait is recorded
  summarize(Count = n()) 

ggplot(data = countsbyrt, aes(x = Count, 
                                     y = reorder(Type, Count))) +
  geom_col(fill = "darkgreen") +
  labs(
       # subtitle = "From Arctic FRED 3.0, 4.0, and 4.1",
       x = "Number of Articles that Record the Root Trait",
       y = "Root Trait Name")
# 
# ggsave(
#   filename = "all_roottraits.png",
#   width = 9,
#   height = 15,
#   dpi = 300
# )
```


